SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2018-06-28 scipy-1.0/paper.tex
ENH: drafted roadmap section of manuscript.

% Packages
\usepackage[super]{nth}
\usepackage{rotating}
\usepackage{makecell}
\usepackage{pifont}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{float}

% New Commands
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

\title {SciPy - some descriptive title here}

\author[1]{Pauli Virtanen}
\author[2,*]{Ralf Gommers}
\author[3]{TODO}
\affil[1]{Affiliation, department, city, postcode, country}
\affil[2]{Affiliation, department, city, postcode, country}
\affil[2]{Affiliation, department, city, postcode, country}

@@

\begin{abstract}
SciPy is an open source scientific computing library for the Python programming language.
SciPy 1.0 was released in late 2017, about 16 years after the original
version 0.1 release. SciPy
has become a \emph{de facto} standard in the community, with more than 600
unique code contributors and millions of downloads per year, despite
the absence of an official project publication, until now. The library
includes functionality spanning clustering, Fourier transforms,
integration, interpolation, file I/O, linear algebra, image processing,
orthogonal distance regression, minimization algorithms, signal processing,
sparse matrix handling, computational geometry, and statistics. In this
work, we highlight some recent technical developments and provide
an overview of the capabilities and development practices of the
SciPy library.
\end{abstract}
\begin{document}

@@

\subsection*{Language preferences}

@@

\subsection*{Polynomial interpolators}

@@

\textbf{Roadmap}

@@

% need some better usage stats sources?
% SciPy stuff from njs blog: https://docs.google.com/spreadsheets/d/1lOLvSF0up4eZyv2ugZi-TM_GIs3gPCkGNIDKXS1Y3w4/edit#gid=406031940
From the user side, there have been at least 15 million downloads
of the SciPy manylinux wheels between the start of 2016 and May of 2018
(njs blog). The recommended manuscript for citing SciPy,
published online 17 years ago, has been cited $>3000$
times. Some of the most prominent citing articles include the IPython
publication, a heavily-cited Nature Biotechnology article, a dark
matter annihlitation study in the field of cosmology, and many of the
other major downstream software libraries in the Python scientific
computing stack.

>>>>

% Packages
\usepackage[super]{nth}
\usepackage{rotating}
\usepackage{makecell}
\usepackage{pifont}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{float}
\usepackage{pgfplots}

\pgfplotsset{compat=newest}

% New Commands
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\code}[1]{\texttt{#1}}

\title{SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python}

\author[1]{Pauli Virtanen}
\author[2,*]{Ralf Gommers}
\author[3,4]{Tyler Reddy}
\author[5]{Anne Archibald}
\author[6]{Andrew Nelson}
\author[7]{Charles Harris}
\author[8]{CJ Carey}
\author[9]{Denis Laxalde}
\author[10]{Eric Larson}
\author[11]{Eric Moore}
\author[12]{Eric Quintero}
\author[13]{Evgeni Burovski}
\author[14]{Jaime Fernández del Río}
\author[15]{Josef Perktold}
\author[16]{Josh Wilson}
\author[17]{Matthew Brett}
\author[18]{Nikolay Mayorov}
\author[19]{Warren Weckesser}
\author[20]{Matt Haberland}
\author[21]{Scott Sievert}
\author[22]{Yu Feng}
\author[23]{Antonio Horta Ribeiro}

\affil[1]{Affiliation, department, city, postcode, country}
\affil[2]{Affiliation, department, city, postcode, country}
\affil[2]{Affiliation, department, city, postcode, country}
\affil[3]{UC Berkeley Institute for Data Science,
          Berkeley, CA, 94720, USA}
\affil[4]{Los Alamos National Laboratory,
	  Theoretical Division 6,
          Los Alamos, NM, 87545, USA}
\affil[5]{Affiliation, department, city, postcode, country}
\affil[6]{Affiliation, department, city, postcode, country}
\affil[7]{Affiliation, department, city, postcode, country}
\affil[8]{Affiliation, department, city, postcode, country}
\affil[9]{Affiliation, department, city, postcode, country}
\affil[10]{Affiliation, department, city, postcode, country}
\affil[11]{Affiliation, department, city, postcode, country}
\affil[12]{Affiliation, department, city, postcode, country}
\affil[13]{Affiliation, department, city, postcode, country}
\affil[14]{Affiliation, department, city, postcode, country}
\affil[15]{Affiliation, department, city, postcode, country}
\affil[16]{Affiliation, department, city, postcode, country}
\affil[17]{Affiliation, department, city, postcode, country}
\affil[18]{Affiliation, department, city, postcode, country}
\affil[19]{Affiliation, department, city, postcode, country}
\affil[20]{Affiliation, department, city, postcode, country}
\affil[21]{Affiliation, department, city, postcode, country}
\affil[22]{Affiliation, department, city, postcode, country}
\affil[23]{Affiliation, department, city, postcode, country}

@@

% NOTE: some usage stats below are from https://libraries.io/pypi/scipy
% as well as github metrics; normally one doesn't put citations directly
% in abstract though (depends on field / journal)
\begin{abstract}
SciPy is an open source scientific computing library for the Python programming language.
SciPy 1.0 was released in late 2017, about 16 years after the original
version 0.1 release. SciPy
has become a \emph{de facto} standard for leveraging scientific algorithms
in the Python programming language, with more than 600
unique code contributors, millions of downloads per year, 161 dependent
packages, and 28700 dependent respositories.
The library includes functionality spanning clustering, Fourier transforms,
integration, interpolation, file I/O, linear algebra, image processing,
orthogonal distance regression, minimization algorithms, signal processing,
sparse matrix handling, computational geometry, and statistics. In this
work, we provide an overview of the capabilities and development practices of the
SciPy library and highlight some recent technical developments.
\end{abstract}
\begin{document}

@@

\subsection*{Language choices}
Python, Cython, Fortran, C and C++ are the programming languages
used to implement scientific algorithms in the SciPy library. An analysis
of our code base using the \texttt{linguist} library provides a detailed
breakdown as \% composition by programming language in SciPy
(Figure~\ref{fig:linguist}).

    \begin{figure}[H]
        \centering

        \begin{tikzpicture}[font=\sffamily]
        \begin{axis}[height=6cm, width=10cm,
        ybar, ymax=60, ymin=0,bar width=1cm,
        ylabel={\% of Scipy code base},
        ytick={0,20,...,60},
        axis lines*=left,
        axis y line shift=0pt,
        enlarge x limits=0.15,
        symbolic x coords={py,f,c,cy,cpp},
        xticklabels={Python,FORTRAN,C,Cython,C++},
        ticklabel style={align=center,font=\sffamily\small,
        /pgf/number format/assume math mode=true},
        xtick=data,
        nodes near coords,
        nodes near coords style={/pgf/number format/assume math mode=true}
        ]
        \addplot coordinates {(py,49.5) (f,25.6) (c,19.5) (cy, 3.0) (cpp,2.3)};
        \end{axis}
        \end{tikzpicture}
        \caption{The breakdown of programming languages used in the
	         SciPy library determined using the linguist library.
		 Small ($<0.5 \%$) amounts of TeX, Matlab, Shell,
		 and Makefile are excluded for clarity and mostly
		 provide supporting roles in tests, building, and
		 documentation.}
        \label{fig:linguist}
    \end{figure}

For implementing new functionality, we have a clear order of language preference.
First Python, if performance is not an issue. If it is, then in order of
preference: Cython, C, C++, Fortran. The main motivation for this is 
maintainability: Cython has the highest abstraction level and most 
Python developers will understand it. C is also widely known,
and easier to deal with for the core development team
than C++ and especially Fortran.

While it is not surprising that Python is heavily used in SciPy, the
usage distribution of the other languages warrants some discussion. Fortran
is an extremely well-established scientific programming language, both
for historical reasons and because of its continued excellent
performance\cite{Koelbel:1993:HPF:562354}. We
wrap the FFTPACK Fortran library for performing Fourier
transforms\cite{SWARZTRAUBER198445, SWARZTRAUBER198251} since
this library has been a standard in the field for 33 years and has a license
that is compatible with our own. Likewise, we wrap the Fortran source
for ODEPACK\cite{citeulike:2644528} as it has been  trusted for the initial 
value problem for ordinary differential equation systems for 30 years. For
similar reasons we also wrap the Fortran libraries QUADPACK\cite{1983qspa.book.....P} (for numerical
integration of one-dimensional functions), FITPACK\cite{Dierckx:1993:CSF:151103} (curve-fitting /
interpolation), ODRPACK\cite{ODRPACK_Boggs} (orthogonal distance regression),
MINPACK\cite{osti_6997568} (minimization of linear and nonlinear equations),
ARPACK\cite{leh:sor:yan96} (solving large scale eigenvalue problems), 
ALGORITHM 644\cite{Amos:1986:APP:7921.214331} (handling Bessel Functions), and 
CDFLIB\cite{CDFLIB_site} (evaluation of cumulative density functions).

Rounding out the top three languages in SciPy is C, which is also extremely
well-established over several decades\cite{Kernighan:1988:CPL:576122} of
scientific computing. Alongside its objected-oriented relative C++, C
can be leveraged in Python after being exposed using the Cython
language. Cython has been described as a creole language that mixes
the best parts of Python and lower-level C / C++
paradigms\cite{behnel2011cython}. We thus often use Cython as a glue
between well-established low-level scientific computing libraries and
the Python interface offered by SciPy. We also use Cython to enable
performance enhancements in Python code, especially for cases where heavily
used inner loops benefit from a compiled code with static typing. Some of the C
libraries that we wrap in SciPy include trlib\cite{doi:10.1080/10556788.2018.1449842} 
(iterative solving of the trust region problem), SuperLU\cite{li05,superlu_ug99} (solution of
large, sparse, nonsymmetric systems of linear equations),
Qhull\cite{Barber:1996:QAC:235815.235821} (computational
geometry), and Cephes\cite{cephes_netlib} (mathematics algorithms).

Therefore, the relative abundance of different programming languages
in the SciPy library results from a combination of the usage of powerful performance
enhancing languages that interact well with Python (i.e., Cython), and
the usage of languages (and their libraries) that have proven reliable
and performant over many decades. The position that SciPy occupies
near the foundation of the scientific Python ecosystem is such that
adoption of new languages or major dependencies is generally unlikely--our
choices are strongly driven by long-term stability. GPU acceleration,
new transpiling libraries, and the latest JIT compilation approaches
(i.e., Numba\cite{Lam:2015:NLP:2833157.2833162}) are very powerful, but currently fall outside the remit
of the main SciPy library. That said, we have recently increased our
efforts to support compatibility with some of these options, and having
our full test suite pass with the PyPy JIT compiler\cite{Bolz:2009:TMP:1565824.1565827}
is now a requirement in our development workflow.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Polynomial interpolators}
\input{poly}

@@

\textbf{Roadmap}
The SciPy Roadmap\cite{SciPy_roadmap} is a continuously-updated document
maintained by the community that describes some of the major directions
for improvement for the project, as well as specific limitations and
matters that require assistance moving forward.

In general, we encourage changes that improvie clarity in the API of the library,
but strongly discourage breaking backwards compatibility, given our position near
the base of the scientific Python computing stack.

Unit test coverage for most recent code additions / adjustments is quite
respectable, but some of the historical components of the code base
would still benefit from increased test coverage. We are still striving
to increase the number of high-quality SciPy usage tutorials, but our
standard documentation for library features is already in excellent shape.

The low-level Cython code in our library (which interacts with C-level
code and exposes it for Python usage) could use some measure of modernization,
including migration to typed memoryviews to handle NumPy arrays in place
of the older syntax.

Some specific submodules of SciPy would benefit from  attention moving forward.
\texttt{fftpack} has some single-precision floating point errors, a caching bug,
and would benefit from the addition of the Bluestein algorithm. There's also
too much community code duplication with \texttt{numpy.fft}, which
requires clarification of best choices and long-term plans.

\texttt{scipy.integrate} now has a new ODE solver interface, and we should aim
to gracefully deprecate the older API. Support for integrating complex-valued
functions and integrating multiple intervals could be beneficial.

\texttt{scipy.interpolate} could benefit from a set of new features including
spline fitting routines with better user control; integration, differentiation,
and arithmetic routines for splines; transparent tensor-product splines,
Non-Uniform Rational B-Splines support; and mesh refinement and coarsening of
B-splines and corresponding tensor products.

\texttt{scipy.io} would benefit from PCM float support, but we otherwise
direct users with more sophisticated needs to domain-specific audio
libraries.

For \texttt{scipy.linalg} we need to remove functions that are duplicate with 
\texttt{numpy.linalg}, \texttt{get\_lapack\_funcs} should always use flapack, and
we should continue to wrap more LAPACK functions. Ideally, we'd also
add type-generic wrappers in the Cython BLAS and LAPACK interfaces, and
transform many of the linear algebra routines into gufuncs.

We're planning to remove \texttt{scipy.misc} as a public module, and
rename \texttt{scipy.odr} to \texttt{scipy.regression} or
\texttt{scipy.fitting}, and include \texttt{optimize.curve\_fit}. This 
module will then provide a home for other fitting functionality.
\texttt{scipy.optimize} would benefit from a few more good global optimizers 
as well as large-scale optimizers.

\texttt{scipy.ndimage} should be completely migrated to a data point
model (values on a grid), instead of a pixel model (elements with centers),
both of which have been used in the past, because the former is now
proving more effective in practice. We also need to standardize the
\texttt{scipy.ndimage.morphology} API.

\texttt{scipy.signal} needs its overlap with \texttt{scipy.ndimage}
to be reduced; B-splines functionality should be moved to \texttt{interpolate},
\texttt{firwin2} can be merged into \texttt{firwin}; Second Order Sections need
to be updated to match the capabilities in other routines, and wavelet-related
routines are somewhat out of place.

\texttt{scipy.sparse} is fairly mature now, but there are many data
formats and the community continues to move forward here so we will aim
to take advantage of their improvements while preserving backwards
compatibility to the extent that it is possible. \texttt{sparse.linalg}
may benefit from wrappers for PROPACK for faster sparse SVD computation.

For \texttt{scipy.spatial}, KDTree will be removed, and cKDTree will be 
renamed to KDTree in a backwards-compatible way. \texttt{distance\_wrap.c} 
likely needs a rewrite in Cython. \texttt{scipy.special} needs precision
improvements for hypergeometric functions, parabolic cylinder functions,
and spheroidal wave functions.

@@

% need some better usage stats sources?
% SciPy stuff from njs blog: https://docs.google.com/spreadsheets/d/1lOLvSF0up4eZyv2ugZi-TM_GIs3gPCkGNIDKXS1Y3w4/edit#gid=406031940
From the user side, there have been at least 15 million downloads
of the SciPy manylinux wheels between the start of 2016 and May of 2018
(njs blog). The recommended source for citing SciPy is simply a reference
to the website / library itself\cite{SciPylib}, and has been cited $>3000$
times. Some of the most prominent citing articles include the IPython
publication\cite{PER-GRA:2007}, a heavily-cited Nature Biotechnology
article\cite{pmid17287757}, a dark
matter annihilitation study in the field of
cosmology\cite{1475-7516-2012-08-007}, and many of the
other major downstream software libraries in the Python scientific
computing stack.
